Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-17 paper.qmd
Initial update for .bib

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data ($k < p$). Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2022], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

These `R` packages were used for the work: `tidyverse` (@hadley2019), `Rtsne` (@jesse2015), `umap` (@tomasz2023), `patchwork` (@thomas2024), `colorspace` (@achim2020), `langevitour` (@harisson2024), `conflicted` (@hadley2023), `reticulate` (@kevin2024), `kableExtra` (@hao2024). These `python` packages were used for the work: `trimap` (@amid2022) and `pacmap` (@yingfan2021). The article was created with `R` packages `rticles` (@jjallaire2024), `knitr` (@yihui2014), and `rmarkdown` (@yihui2020). The project's GitHub repository ([https://github.com/JayaniLakshika/paper-nldr-vis-algorithm](https://github.com/JayaniLakshika/paper-nldr-vis-algorithm)) contains all materials required to reproduce this article.

>>>>

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data ($k < p$). Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

These `R` packages were used for the work: `tidyverse` (@hadley2019), `Rtsne` (@jesse2015), `umap` (@tomasz2023), `patchwork` (@thomas2024), `colorspace` (@achim2020), `langevitour` (@harisson2024), `conflicted` (@hadley2023), `reticulate` (@kevin2024), `kableExtra` (@hao2024). These `python` packages were used for the work: `trimap` (@amid2019) and `pacmap` (@yingfan2021). The article was created with `R` packages `rticles` (@jjallaire2024), `knitr` (@yihui2014), and `rmarkdown` (@yihui2020). The project's GitHub repository ([https://github.com/JayaniLakshika/paper-nldr-vis-algorithm](https://github.com/JayaniLakshika/paper-nldr-vis-algorithm)) contains all materials required to reproduce this article.
